In-silico molecular enrichment and clearance of the human intracranial space

### 2025-01-26 paper/main.tex
Updates from Overleaf

Our periarterial network extends from the internal carotid arteries
and basilar artery through up to 18 bifurcations to reach upstream
network ends located within the SAS or up to 6 mm inside the
parenchyma (\Cref{fig:results1}A--C, \Cref{fig:pvs}A) and includes
major surface arteries of radius 0.5--1.4 mm. Imposing the pressure
field induced by CSF production at the periarterial network ends
induces ultraslow steady CSF flow of variable direction in these PVSs,
with an average velocity of 0.08 \textmu m/s (antegrade) and a maximum
velocity of 0.55 \textmu m/s (\Cref{fig:pvs}B). On the other hand,
traveling waves of arterial wall motion (\Cref{fig:pvs}C), such as the
pulse wave or other vasomotion~\cite{vanveluw2020vasomotion,
  munting2023spontaneous, bojarskaite2023sleep, broggini2024long,hauglund2025norepinephrine},
also induce net directional flow in the
PVS~\cite{kedarasetti2020arterial, kedarasetti2020functional,
  coenen2021lubrication, gjerde2023directional, nozaleda2024arterial}
-- of magnitude and direction depending on the amplitude, frequency,
and length of the waves and the characteristics of the perivascular
network~\cite{gjerde2023directional}. Applying a
semi-analytic model of the net flow induced by peristalsis in
perivascular networks~\cite{gjerde2023directional} (see Methods), we
estimate that the cardiac pulse wave alone, traveling at a frequency
of 1 Hz with a wavelength of 2.0 m and a 1\% wall
displacement~\cite{jung2021novel}, will induce mainly antegrade PVS
flow with an average net velocity of 0.92 \textmu m/s while reaching up
to 7.31 \textmu m/s near larger, ventral vessels such as the MCA
(\Cref{fig:pvs}D). The same theory predicts that strong ultraslow
vasomotions, if traveling antegrade at 0.1Hz with a wavelength of 0.02
m and a 10\% displacement~\cite{broggini2024long}, will induce both
retrograde and antegrade net PVS flow with an average velocity of
13.05 \textmu m/s and maximum velocity 54.44 \textmu m/s
(\Cref{fig:pvs}E).

We study the effect of such rapid PVS flow on intracranial molecular
enrichment by comparing the more conservative baseline model and a
high PVS flow model, where the former still includes the net flow
contributions from CSF production and the cardiac pulse wave, while
the latter additionally includes the ultraslow vasomotion
contribution. In both models, tracers were first observed at the basal
artery after 48 minutes (\Cref{fig:pvs}F), defined by their first-time
arrival (FTA), the time at which the concentration first exceeds $0.1$
mmol/l. At all upstream locations along the middle cerebral arteries
(MCAs) and anterior cerebral artery (ACA), we found substantially
reduced FTAs with higher PVS flow (\Cref{fig:pvs}G), up to 2.5 hours
earlier in the left M2 segment of the MCA. PVS concentrations peaked
before the surrounding tissue (0:48h vs 1:12h at the MCA and 1:00h vs
1:48h at MCA2) with high PVS flow, while these peaks occurred nearly
simultaneously in the baseline model. The earlier appearance and the
delay in time-to-peak between the PVS and surrounding tissue clearly
indicates the directionality of tracer enrichment -- it first arrives
in the PVS and subsequently spreads into the tissue and SAS
(\Cref{fig:pvs}H, I), and especially so with higher PVS flow. These
observations also hold on an aggregated level: computing the total
amount of tracer in the PVS as a function of the distance to the
arterial network roots, we find accelerated tracer transport with
higher PVS flow, especially at earlier time points (2--9 hours)
(\Cref{fig:pvs}J). Moreover, differences in tracer transport along the
PVS translate to altered enrichment patterns on the whole-organ
scale. For instance, after 4-6 hours, the faster-moving tracer in the
PVS is clearly visible in the space adjacent to the PVS of the ACA, MCA, and other arteries (\Cref{fig:pvs}L). In contrast,
there are no clear signs of early enrichment surrounding the PVS in
the baseline model (\Cref{fig:pvs}K).

A natural question is whether early PVS enrichment could be the result
of increased dispersion rather than net flow in the
PVS\cite{asgari2016glymphatic,sharp2019dispersion,bojarskaite2023sleep,asgari2016glymphatic,troyetsky2021dispersion}. Interestingly,
even increasing the dispersion factor by $100 \times$ in the PVS,
induced only minor changes in the global spreading rate compared to the
baseline (\Cref{fig:pvs}G, M), thus indicating a negative answer to
this question.
\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure3.png}
    \caption{\textbf{Perivascular flow shapes and accelerates molecular enrichment.}}
    \label{fig:pvs}
\end{figure}
\begin{figure}
\ContinuedFloat
  \caption{\textbf{(cont.)}
    A) Surface arterial network colored by arterial radius with vessel locations labeled;
    B) Net PVS flow induced by CSF production (cf.~\Cref{fig:csf}D) -- histograms show the relative frequencies of local periarterial velocities. Positive values indicate antegrade PVS flow, while negative values indicate retrograde PVS flow;
    C) Illustration of the concept of peristaltic pumping;
    D) Estimated net PVS flow induced by pulse wave peristaltic pumping;
    E) Estimated net PVS flow induced by vasomotion/slow wave peristaltic pumping;
    F) Comparison of mean tracer concentration in the PVS and over the outer PVS surface at key locations of the arterial tree, for the baseline (lower) and high PVS flow (upper) models, with labels: ``peak'': time-to-peak (h), ``$\Delta t$'': time difference between peaks (h), and ``FTA'': first-time tracer arrival (h).
    G) first-time arrival for the main trunks of the periarterial network over the distance from the closest network root node (BA, ICA-R or ICA-L for the baseline, the high PVS flow and the high PVS dispersion models);
    H, I) 2D slices showing zoom-in on the region surrounding the MCA-R and MCA2-R for the baseline (upper) and high PVS flow (lower) models. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF));
    J) Total amount of tracer in the periarterial spaces as a function of the distance to the nearest network root over time for the baseline (left) and the high PVS flow (right) models;
    K--M) sagittal (upper) and coronal (lower) views of the simulated tracer concentrations overlayed on the T1-weighted MR image after 4 and 6h for the baseline (K), high PVS flow (L) and high PVS dispersion (M) models (concentration opacity linearly increasing with concentration value, pial surface outlined in pink, arachnoid membrane in cyan, and arteries in dark red).}
\end{figure}

@@

From T1-weighted MR images of a human subject~\cite{hodneland2019new},
we first automatically segment the brain parenchyma and CSF spaces,
including the ventricular system and SAS, using
Synthseg~\cite{billot2023robust,billot2023synthseg}
(\Cref{fig:results1}A--B). We next manually adjust the segmentation to
accurately represent the connections between CSF spaces (aqueduct,
median aperture, tentorium cerebelli), smoothen, and finally extract
surface representations of the outer (arachnoid) boundary, the pial
membrane and other interfaces. Conforming to these surface and
interface representations, we generate a tetrahedral mesh $\Omega$
representing the complete intracranial volume as the union of the
parenchyma $\Omega_{\rm PAR}$ and CSF spaces $\Omega_{\rm CSF}$, using
fTetWild~\cite{hu2020fast} (\Cref{fig:csf}A--B). The resulting
computational mesh consists of $233\,592$ vertices and $1\,290\,131$
mesh cells, which vary between $0.22$ mm and $8.9$ mm in mesh cell
diameter. The parenchyma has a total volume of $1318$ ml, whereas the
ventricles and the outer SAS contribute $32$ ml and $329$ ml,
respectively, to a total CSF volume of $361$ml, in agreement with
recent estimates~\cite{hladky2024regulation}.

@@

To represent surface networks of periarterial and perivenous spaces,
we use Kiminaro~\cite{william_silversmith_2021_5539913} to separately
skeletonize time-of-flight angiography (ToF) and quantitative
susceptibility mapping (QSM) images from the same human
subject~\cite{hodneland2019new}. This technique yields networks of
one-dimensional curves, each curve $\Lambda^i$ indicating the
centerline of a blood vessel segment and labeled with its lumen radius
$R_1^i$ (\Cref{fig:results1}C, \Cref{fig:pvs}A). We also assign an
outer radius $R_2^i > R_1^i$ to each segment $i$, form the annular
cylinder ensheathing $\Lambda^i$ and define the union of these as the
PVS. The periarterial network graph consists of $12\,708$ edges
connected at $12\,562$ inner nodes and with $147$ end nodes, and the
associated domain is denoted by $\Lambda_a$. We identify three of the
end nodes -- corresponding to the two internal carotid arteries (ICAs)
and the basilar artery (BA) -- as root nodes, and designate the other
ends as leaf nodes. The resulting perivenous network graph consists of
$\fixme{N}$ edges connected at $\fixme{N}$ inner nodes and with
$\fixme{N}$ end nodes, and the associated domain is denoted by
$\Lambda_v$. \discuss{Finally, we label the outer surface of the PVSs
  associated with $\Lambda_a$ and $\Lambda_v$, by $\Gamma_a$ and
  $\Gamma_v$ respectively.}

@@

Using the theoretical framework introduced by Gjerde et
al.~\cite{gjerde2023directional}, we also compute analytic estimates
for the time-average perivascular flow rate $\langle Q' \rangle$
induced by peristalsis in the arterial network $\Lambda_a$; i.e.~the
net flow induced by traveling waves of arterial wall motion of
frequency $f$, amplitude $\varepsilon$ and wave lengths $\lambda$
(\Cref{fig:pvs}D). The corresponding contribution to the periarterial
flow velocity is defined for each segment $\Lambda_i$ as $\langle Q'_i
\rangle/A_i$ where $A_i = \pi (R_2^i - R_1^i)$ is the cross-section
area of the PVS segment and $\langle Q_i' \rangle$ is the estimated
mean flow rate (see~\ref{sec:app:peristalsis} for more details). Two
wall motion patterns are considered, yielding $\hat{u}_a^c$ and
$\hat{u}_a^v$ respectively: \emph{cardiac} pulsations with $f =
1.0$Hz, $\lambda = 2$ m and $\varepsilon = 1\%$ (\Cref{fig:pvs}E), and
very low-frequency \emph{vasomotion} with $f = 0.1$Hz, $\lambda =
0.02$m, and $\varepsilon = 10\%$ (\Cref{fig:pvs}F). No flow induced by
peristalsis is considered for the perivenous network. In the baseline
model, we combine the contributions from CSF production and cardiac
peristalsis to set $\hat{u}$ in~\eqref{eq:multi_transport} as
$\hat{u}_a = \hat{u}_a^p + \hat{u}_a^c$ in $\Lambda_a$, and $\hat{u}_v
= \hat{u}_v^p$ in $\Lambda_v$.

>>>>

Our periarterial network extends from the internal carotid arteries
and basilar artery through up to 18 bifurcations to reach upstream
network ends located within the SAS or up to 6 mm inside the
parenchyma (\Cref{fig:results1}A--C, \Cref{fig:pvs_a}A) and includes
major surface arteries of radius 0.5--1.4 mm. Imposing the pressure
field induced by CSF production at the periarterial network ends
induces ultraslow steady CSF flow of variable direction in these PVSs,
with an average velocity of 0.08 \textmu m/s (antegrade) and a maximum
velocity of 0.55 \textmu m/s (\Cref{fig:pvs_a}B). On the other hand,
traveling waves of arterial wall motion (\Cref{fig:pvs_a}C), such as the pulse wave or other vasomotion~\cite{vanveluw2020vasomotion,
  munting2023spontaneous, bojarskaite2023sleep, broggini2024long,hauglund2025norepinephrine},
also induce net directional flow in the
PVS~\cite{kedarasetti2020arterial, kedarasetti2020functional,
  coenen2021lubrication, gjerde2023directional, nozaleda2024arterial}
-- of magnitude and direction depending on the amplitude, frequency,
and length of the waves and the characteristics of the perivascular
network~\cite{gjerde2023directional}. Applying a
semi-analytic model of the net flow induced by peristalsis in
perivascular networks~\cite{gjerde2023directional} (see Methods), we
estimate that the cardiac pulse wave alone, traveling at a frequency
of 1 Hz with a wavelength of 2.0 m and a 1\% wall
displacement~\cite{jung2021novel}, will induce mainly antegrade PVS
flow with an average net velocity of 0.92 \textmu m/s while reaching up
to 7.31 \textmu m/s near larger, ventral vessels such as the MCA
(\Cref{fig:pvs_a}D). The same theory predicts that strong ultraslow
vasomotions, if traveling antegrade at 0.1Hz with a wavelength of 0.02
m and a 10\% displacement~\cite{broggini2024long}, will induce both
retrograde and antegrade net PVS flow with an average velocity of
13.05 \textmu m/s and maximum velocity 54.44 \textmu m/s
(\Cref{fig:pvs_a}E).

We study the effect of such rapid PVS flow on intracranial molecular
enrichment by comparing the more conservative baseline model and a
high PVS flow model, where the former still includes the net flow
contributions from CSF production and the cardiac pulse wave, while
the latter additionally includes the ultraslow vasomotion
contribution. In both models, tracers were first observed at the basal
artery after 48 minutes (\Cref{fig:pvs_b}A), defined by their first-time arrival (FTA), the time at which the concentration first exceeds $0.1$ mmol/l. At all upstream locations along the middle cerebral arteries
(MCAs) and anterior cerebral artery (ACA), we found substantially
reduced FTAs with higher PVS flow (\Cref{fig:pvs_a}B), up to 2.5 hours
earlier in the left M2 segment of the MCA. PVS concentrations peaked
before the surrounding tissue (0:48h vs 1:12h at the MCA and 1:00h vs
1:48h at MCA2) with high PVS flow, while these peaks occurred nearly
simultaneously in the baseline model. The earlier appearance and the
delay in time-to-peak between the PVS and surrounding tissue clearly
indicates the directionality of tracer enrichment -- it first arrives
in the PVS and subsequently spreads into the tissue and SAS
(\Cref{fig:pvs_a}C, D), and especially so with higher PVS flow. These
observations also hold on an aggregated level: computing the total
amount of tracer in the PVS as a function of the distance to the
arterial network roots, we find accelerated tracer transport with
higher PVS flow, especially at earlier time points (2--9 hours)
(\Cref{fig:pvs_b}E). Moreover, differences in tracer transport along the
PVS translate to altered enrichment patterns on the whole-organ
scale. For instance, after 4-6 hours, the faster-moving tracer in the
PVS is clearly visible in the space adjacent to the PVS of the ACA, MCA, and other arteries (\Cref{fig:pvs_b}G). In contrast,
there are no clear signs of early enrichment surrounding the PVS in
the baseline model (\Cref{fig:pvs_b}F).

A natural question is whether early PVS enrichment could be the result
of increased dispersion rather than net flow in the
PVS\cite{asgari2016glymphatic,sharp2019dispersion,bojarskaite2023sleep,asgari2016glymphatic,troyetsky2021dispersion}. Interestingly,
even increasing the dispersion factor by $100 \times$ in the PVS,
induced only minor changes in the global spreading rate compared to the
baseline (\Cref{fig:pvs_b}F, H), thus indicating a negative answer to
this question.

\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure3a.png}
    \caption{\textbf{A static pressure gradient and peristaltic pumping are potential drivers of perivascular flow.}
    A) Surface arterial network colored by arterial radius with vessel locations labeled;
    B) Net PVS flow induced by CSF production (cf.~\Cref{fig:csf}D) -- histograms show the relative frequencies of local periarterial velocities. Positive values indicate antegrade PVS flow, while negative values indicate retrograde PVS flow;
    C) Illustration of the concept of peristaltic pumping;
    D) Estimated net PVS flow induced by pulse wave peristaltic pumping;
    E) Estimated net PVS flow induced by vasomotion/slow wave peristaltic pumping.}
    \label{fig:pvs_a}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.99\textwidth]{figures/figure3b.png}
    \caption{\textbf{Perivascular flow shapes and accelerates molecular enrichment.}
    A) Comparison of mean tracer concentration in the PVS and over the outer PVS surface at key locations of the arterial tree, for the baseline (lower) and high PVS flow (upper) models, with labels: ``peak'': time-to-peak (h), ``$\Delta t$'': time difference between peaks (h), and ``FTA'': first-time tracer arrival (h).
    B) first-time arrival for the main trunks of the periarterial network over the distance from the closest network root node (BA, ICA-R or ICA-L for the baseline, the high PVS flow and the high PVS dispersion models);}
    \label{fig:pvs_b}
\end{figure}
\begin{figure}
\ContinuedFloat
\caption{(\textbf{cont.})
    C, D) 2D slices showing zoom-in on the region surrounding the MCA-R and MCA2-R for the baseline (upper) and high PVS flow (lower) models. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF));  
    E) Total amount of tracer in the periarterial spaces as a function of the distance to the nearest network root over time for the baseline (left) and the high PVS flow (right) models;
    F--H) sagittal (upper) and coronal (lower) views of the simulated tracer concentrations overlayed on the T1-weighted MR image after 4 and 6h for the baseline (K), high PVS flow (L) and high PVS dispersion (M) models (concentration opacity linearly increasing with concentration value, pial surface outlined in pink, arachnoid membrane in cyan, and arteries in dark red).}
\end{figure}

@@

From T1-weighted MR images of a 26-year-old, healthy male volunteer~\cite{hodneland2019new},
we first automatically segment the brain parenchyma and CSF spaces,
including the ventricular system and SAS, using
Synthseg~\cite{billot2023robust,billot2023synthseg}
(\Cref{fig:results1}A--B). We next manually adjust the segmentation to
accurately represent the connections between CSF spaces (aqueduct,
median aperture, tentorium cerebelli), smoothen, and finally extract
surface representations of the outer (arachnoid) boundary, the pial
membrane and other interfaces. Conforming to these surface and
interface representations, we generate a tetrahedral mesh $\Omega$
representing the complete intracranial volume as the union of the
parenchyma $\Omega_{\rm PAR}$ and CSF spaces $\Omega_{\rm CSF}$, using
fTetWild~\cite{hu2020fast} (\Cref{fig:csf}A--B). The resulting
computational mesh consists of $233\,592$ vertices and $1\,290\,131$
mesh cells, which vary between $0.22$ mm and $8.9$ mm in mesh cell
diameter. The parenchyma has a total volume of $1318$ ml, whereas the
ventricles and the outer SAS contribute $32$ ml and $329$ ml,
respectively, to a total CSF volume of $361$ml, in agreement with
recent estimates~\cite{hladky2024regulation}.

@@

To represent surface networks of periarterial and perivenous spaces,
we use Kiminaro~\cite{william_silversmith_2021_5539913} to separately
skeletonize time-of-flight angiography (ToF) and quantitative
susceptibility mapping (QSM) images from the same human
subject~\cite{hodneland2019new}. This technique yields networks of
one-dimensional curves, each curve $\Lambda^i$ indicating the
centerline of a blood vessel segment and labeled with its lumen radius
$R_1^i$ (\Cref{fig:results1}C, \Cref{fig:pvs_a}A). We also assign an
outer radius $R_2^i > R_1^i$ to each segment $i$, form the annular
cylinder ensheathing $\Lambda^i$ and define the union of these as the
PVS. The periarterial network graph consists of $12\,708$ edges
connected at $12\,562$ inner nodes and with $147$ end nodes, and the
associated domain is denoted by $\Lambda_a$. We identify three of the
end nodes -- corresponding to the two internal carotid arteries (ICAs)
and the basilar artery (BA) -- as root nodes, and designate the other
ends as leaf nodes. The resulting perivenous network graph consists of
$\fixme{N}$ edges connected at $\fixme{N}$ inner nodes and with
$\fixme{N}$ end nodes, and the associated domain is denoted by
$\Lambda_v$. \discuss{Finally, we label the outer surface of the PVSs
  associated with $\Lambda_a$ and $\Lambda_v$, by $\Gamma_a$ and
  $\Gamma_v$ respectively.}

@@

Using the theoretical framework introduced by Gjerde et
al.~\cite{gjerde2023directional}, we also compute analytic estimates
for the time-average perivascular flow rate $\langle Q' \rangle$
induced by peristalsis in the arterial network $\Lambda_a$; i.e.~the
net flow induced by traveling waves of arterial wall motion of
frequency $f$, amplitude $\varepsilon$ and wave lengths $\lambda$
(\Cref{fig:pvs_a}C). The corresponding contribution to the periarterial
flow velocity is defined for each segment $\Lambda_i$ as $\langle Q'_i
\rangle/A_i$ where $A_i = \pi (R_2^i - R_1^i)$ is the cross-section
area of the PVS segment and $\langle Q_i' \rangle$ is the estimated
mean flow rate (see~\ref{sec:app:peristalsis} for more details). Two
wall motion patterns are considered, yielding $\hat{u}_a^c$ and
$\hat{u}_a^v$ respectively: \emph{cardiac} pulsations with $f =
1.0$Hz, $\lambda = 2$ m and $\varepsilon = 1\%$ (\Cref{fig:pvs_a}D), and very low-frequency \emph{vasomotion} with $f = 0.1$Hz, $\lambda =
0.02$m, and $\varepsilon = 10\%$ (\Cref{fig:pvs_a}E). No flow induced by peristalsis is considered for the perivenous network. In the baseline model, we combine the contributions from CSF production and cardiac peristalsis to set $\hat{u}$ in~\eqref{eq:multi_transport} as
$\hat{u}_a = \hat{u}_a^p + \hat{u}_a^c$ in $\Lambda_a$, and $\hat{u}_v
= \hat{u}_v^p$ in $\Lambda_v$.
